Urban Diversity Through an Information-Theoretic Lens

### 2016-08-31 tex/paper/methods.tex
minor updates to slides and paper

	Finally, we consider the local information $J(X,Y)$ and its connection to spatial exposure. The authors of \cite{Reardon2002} define \emph{spatial exposure} as ``the extent that members of one group encounter members of another group...in their local spatial environments'' (page 126). Our contention is that $J(X,Y)$ measures exposure conditional on a fixed level of evenness (as measured by $I(X,Y)$). To see this, consider cities (c) and (d) in Figure \ref{fig:toy}. These cities are equally diverse as measured by $H(Y)$ and equally uneven as measured by $I(X,Y)$. The difference between them is that (c) consists of two large, racially-distinct neighborhoods, of which only residents near the center are exposed to residents of the other. On the other hand, in (d), all residents are just one ``block'' away from residents who do not share their race. $J(X,Y)$ detects this difference because the localized mutual information $I_r$ in terms of which $J(X,Y)$ is defined is precisely the unevenness in a small area. We should emphasize, however, that $J(X,Y)$ should always be considered in the context of $I(X,Y)$. City (d) clearly has more exposure than (c), but less than the perfectly even (b). The difference is that (b) has no neighborhood structure at all.

@@

\subsection*{Desirable Properties of Information Measures}
	How do these information measures There has also been much work showing the desirable properties of various segregation indicies. To give a brief sampling, indices should be invariant to changes in overall population size; they should decrease when populations ``even themselves out,'' and they should behave predictably under aggregation. Since we are presenting $I(X,Y)$ and $J(X,Y)$ as a suite of complementary information measures, we consider each in turn. The author of \cite{Roberto2015a} shows that the mutual information $I(X,Y)$ (which she calls the ``Divergence Index'') satisfies these and other desirable properties, generally as well or better than existing alternatives. We highlight one property of $I(X,Y)$ for special note, as this property will be central to our development of natural neighborhood identification. The principle of ``additive decomposability'' \cite{Reardon2002} stipulates that, when the data is grouped along either racial or spatial axes, a good segregation index should split into ``within group'' and ``between group'' components. In the context of the mutual information $I(X,Y)$, additive decomposability is simply the familiar chain rule of mutual information. For concreteness, let $C$ be a random variable giving the cluster label of location $X$; importantly, $C$ is completely determined by $X$. Then, the chain rule expresses additive decomposability as 
	\begin{equation}
		I(X,Y) = I(C,Y) + I(X,Y|C)\;; \label{eq:information_decomp}
	\end{equation}
	i.e. the information I have about $Y$ given that I know $X$ is equal to the information I have if I first learn the cluster $C$, plus the amount of additional information I gain if I subsequently learn the exact location $X$ as well. The first term is interpretable as the between-group information, while the second is interpretable as the within-group information. It is therefore comparable to various ``sum of squares'' decompositions that frequently appear in classical statistics. Indeed, it is possible to show that, when the differences between locations are small, the mutual information can be interpreted as a variance, in which case \eqref{eq:information_decomp} expresses the sum of squares composition directly. A similar version of the chain rule expresses additive decomposability for aggregation of racial groups rather than spatial locations.

\color{gray} % this section is very WIP.

	What of $J(X,Y)$? Below, we provide a brief overview of the properties of $J(X,Y)$. $J(X,Y)$ possesses a variety of intuitive properties, and those that it does not possess either underscore the importance of reading it in conjunction with $I(X,Y)$ or do not apply to explicitly spatial measures.  See \cite{Reardon2002,Reardon2004} for further discussion of these criteria. The mean local information $J(X,Y)$ satisfies: 

@@

		\item[Organizational Equivalence:] Both the theoretical definition \eqref{eq:def_J} or the procedure to compute it from tract data remain unchanged when a tract is subdivided into smaller tracts, each of which with identical demographic structure. 

@@

		\item[Scale Interpretability:] We have $J(X,Y) = 0$ if and only if $p(Y|X)$ is constant on each connect component of the metric space $M$. When only one connected component exists, this implies that every locale has the same demographic structure as the global environment. When multiple connected components exist (e.g. the city is divided by a river), demographics must be constant in space on each side of the division. $J(X,Y)$ does not satisfy the additional condition of achieving its maximum value when all locales are monoracial, but this point only emphasizes that $J(X,Y)$ and $I(X,Y)$ should be read jointly. When all locales are monoracial, $I(X,Y)$ achieves its maximum value and $J(X,Y) = 0$. 
		\item[Boundary Independent:] $J(X,Y)$ is defined in terms of a continuous underlying probability distribution $p(X,Y)$, rather than arbitrarily-defined tracts. In computational practice, $J(X,Y)$ is indeed sensitive to the boundaries supplied with the data; however, \eqref{eq:approx} guarantees that this sensitivity vanishes as the resolution grows sufficiently small.  
		\item[Exchanges:] Exchanges that tend to ``smooth out'' demographic distributions in space reduce both $I(X,Y)$ and $J(X,Y)$.  

@@

	The mean local information $J(X,Y)$ fails to one of the criteria enumerated in \cite{Reardon2002,Reardon2004}; that of \textbf{additive spatial decomposability}. This criterion states that ``If X spatial subareas are aggregated into Y larger spatial areas, then a segregation measure should be decomposable into a sum of within- and between-area components'' \cite{Reardon2004}, page 136. However, considering that we have imposed additional spatial structure on our model, this criterion does not appear to be motivated. Consider, for example, Figure \ref{fig:decomposability}. Aggregating the two central regions transforms (e) into (f), erasing all spatial variability. Any measure should therefore have a between-groups component equal to 0. On the other hand, the within-groups component can only consider the middle white/black boundary between the central two regions. The sum of the between-groups and within-groups components must therefore consist only in information included in the middle boundary. However, the spatial variability in city (e) is not exhausted by this middle boundary; there are a total of six other frontiers of racial difference that should be considered in any spatial segregation measure. We therefore content that additive decomposability is a desideratum of nonspatial measures (it is satisfied by $I(X,Y)$), but not of explicitly spatial ones. 
	\begin{figure}
		\centering
		\includegraphics[width=.5\textwidth]{figs/decomposability.png}
		\caption{A problem with additive spatial decomposability: aggregating the inner two regions transforms (e) into (f), smoothing out all spatial variability. A ``between groups'' term must therefore vanish, while a ``within-groups'' term considers only the inner boundary; not the outer ones.}
		\label{fig:decomposability}
	\end{figure}
	We have not exhausted the full complement of desiderata for segregation measures. However, many of the remaining ones -- such as composition invariance -- are either vague or controversial, and we will not consider further. 
\color{black}

@@

	Equation \eqref{eq:information_decomp} motivates a simple scheme for identifying natural neighborhoods based on maximizing the between-group term $I(C,Y)$. If all I had access to were the cluster labels $C$ and not the locations $X$, then the information I had about $Y$would be $I(C,Y)$. A ``good'' clustering therefore maximizes the between-cluster information $I(Y,C)$, which entails minimizing the within-cluster information $I(X,Y|C)$. Solving this problem exactly is a challenging discrete optimization problem, and may not be computationally tractable. We can, however, construct a greedy algorithm which leads to satisfactory results. Suppose that we face the problem of choosing a pair of tracts $\{i*,j*\}$ to cluster together. The reduction in information associated with aggregating the locations $I$ into a single cluster is 
	\begin{align}
		d(i,j) &\triangleq  \sum_{k \in \{i,j\}} p(X = k)D[p(Y|X = k)\| p(Y|X \in \{i,j\})] \\
		&- p(X\in \{i,j\})D[p(Y|X\in \{i,j\}) \| p(Y)] \label{eq:info_dist}
	\end{align}
	where $p(Y) = \sum_{x} p(x,Y)$ is the global marginal distribution. the first term is the information associated with the two separate tracts, while the second is the information associated with a merged tract. Equation \eqref{eq:info_dist} defines a natural information distance between locations $i$ and $j$. Like the KL divergence, this distance is strictly nonnegative; unlike the KL divergence, it is symmetric, and defines an axiomatic metric on the space of tracts. Importantly, we can therefore use the distance $d(i,j)$ as a dissimilarity measure for the purposes of clustering. Our greedy procedure is simple: at each iteration, determine
	\begin{equation}
		(i^*,j^*) = \argmin_{i \text{ neighbors } j} d(i,j),
	\end{equation}
	and then combine $i^*$ amd $j^*$ into a single tract, repeating until only one cluster remains. This procedure defines a form of agglomerative hierarchical clustering distinguished by two characteristics: its spatial constraints and its pursuit of minimal information loss at each step. As a greedy algorith, it possesses no guarantees for optimal solutions, but in practice its performance leads to intuitive, racially-coherent regions. It thereby enables a study of spatial difference using non-arbitrarily-defined regions.

>>>>

	Finally, we consider the local information $J(X,Y)$ and its connection to spatial exposure. The authors of \cite{Reardon2002} define \emph{spatial exposure} as ``the extent that members of one group encounter members of another group...in their local spatial environments'' (page 126). Our contention is that $J(X,Y)$ measures exposure conditional on a fixed level of evenness (as measured by $I(X,Y)$). To see this, consider cities (c) and (d) in Figure \ref{fig:toy}. These cities are equally diverse as measured by $H(Y)$ and equally uneven as measured by $I(X,Y)$. The difference between them is that (c) consists of two large, racially-distinct neighborhoods, of which only residents near the center are exposed to residents of the other. On the other hand, in (d), all residents are just one ``block'' away from residents who do not share their race. $J(X,Y)$ detects this difference because the localized mutual information $I_r$ in terms of which $J(X,Y)$ is defined precisely as the unevenness in a small area. We should emphasize, however, that $J(X,Y)$ should always be considered in the context of $I(X,Y)$. City (d) clearly has more exposure than (c), but less than the perfectly even (b). The difference is that (b) has no neighborhood structure at all.

@@

\subsection*{Properties of Information Diversity Measures}
	There has been much work showing the desirable properties of various segregation indices. To give a brief sampling, many scholars agree that such indices should be invariant to changes in overall population size; they should decrease when populations ``even themselves out,'' and they should behave predictably under aggregation. How do the measures $I(X,Y)$ and $J(X,Y)$ compare? The author of \cite{Roberto2015a} shows that the mutual information $I(X,Y)$ (which she calls the ``Divergence Index'') satisfies the core desiderata of segregation indices, generally as well or better than existing alternatives. We highlight one property of $I(X,Y)$ for special note, as it will be central to our development of natural neighborhood identification. The principle of ``additive decomposability'' \cite{Reardon2002} stipulates that, when the data is grouped along either racial or spatial axes, a good segregation index should split into ``within group'' and ``between group'' components. In the context of the mutual information $I(X,Y)$, additive decomposability is simply the familiar chain rule of mutual information. For concreteness, let $C$ be a random variable giving the cluster label of location $X$; importantly, $C$ is completely determined by $X$. Then, the chain rule expresses additive decomposability as 
	\begin{equation}
		I(X,Y) = I(C,Y) + I(X,Y|C)\;; \label{eq:information_decomp}
	\end{equation}
	This information identity has a simple epistemic interpretation: he information I have about someone's race $Y$ given that I know their residence $X$ is equal to the information I have if I first learn their broad neighborhood $C$, plus the amount of additional information I gain if I subsequently learn the exact residence $X$ as well. Mathematically, this expression decomposes $I(X,Y)$ into two useful components. The first term is interpretable as the between-group information, while the second is interpretable as the within-group information. It is therefore comparable to the familiar between-group/within-group ``sum of squares'' decompositions that frequently appear in classical statistical analysis. A similar version of the chain rule expresses additive decomposability for aggregation of racial groups rather than spatial locations. In the next subsection, we use \eqref{eq:information_decomp} to define a clustering procedure to identify natural neighborhoods in cities.

	We now turn to the properties of $J(X,Y)$. Our core argument is that $J(X,Y)$ possesses a variety of intuitive properties, and that those it does not either underscore the importance of reading it in conjunction with $I(X,Y)$ or need not apply to spatialized measures.  See \cite{Reardon2002,Reardon2004} for detailed discussion of these criteria. 

@@

		\item[Organizational Equivalence:] Both the theoretical definition \eqref{eq:def_J} and the procedure to compute it from tract data remain unchanged when a tract is subdivided into smaller tracts, each of which with identical demographic structure. 

@@

		\item[Scale Interpretability:] We have $J(X,Y) = 0$ if and only if $p(Y|X)$ is constant on each connected component of the metric space $M$. When only one connected component exists, this implies that every locale has the same demographic structure as the global environment. When multiple connected components exist (e.g. the city is divided by a river), demographics must be constant in space on each side of the division. $J(X,Y)$ does not satisfy the additional condition of achieving its maximum value when all locales are monoracial, but this point only emphasizes that $J(X,Y)$ and $I(X,Y)$ should be read jointly. When all locales are monoracial, $I(X,Y)$ achieves its maximum value and $J(X,Y) = 0$. 
		\item[Boundary Independence:] $J(X,Y)$ is defined in terms of a continuous underlying probability distribution $p(X,Y)$, rather than arbitrarily-defined tracts. In computational practice, $J(X,Y)$ is indeed sensitive to the boundaries supplied with the data; however, \eqref{eq:approx} guarantees that this sensitivity vanishes as the resolution grows sufficiently small. No alternative measure can overcome limited data resolution, and, to our knowledge, no other measure comes with the same assurance of resolution-independence in the limit. 
		\item[Transfers and Exchanges:] The core idea of the principles of transfers and exchanges is that population permutations that tend to ``smooth out'' differences should reduce segregation measures. However, the concrete principles as stated in \cite{Reardon2002} do not hold of $J(X,Y)$. The core point is that $J(X,Y)$ compares distributions to those nearby, so whether or not a transfer or exchange reduces $J(X,Y)$depends not only on the two regions involved in the transfer or exchange, but also the regions surrounding them. $I(X,Y)$ does satisfy the two principles. % Consider trying to demonstrate the spatial transfer principle for the computational version. 
		\item[Additive Spatial Decomposability:] This criterion states that ``If X spatial subareas are aggregated into Y larger spatial areas, then a segregation measure should be decomposable into a sum of within- and between-area components'' \cite{Reardon2004}, page 136. However, considering that we have imposed additional spatial structure on our model, this criterion does not appear to be motivated. Consider, for example, Figure \ref{fig:decomposability}. Aggregating the two central regions transforms (e) into (f), erasing all spatial variability. Any measure should therefore have a between-groups component equal to 0. On the other hand, the within-groups component can only consider the middle white/black boundary between the central two regions. The sum of the between-groups and within-groups components must therefore consist only in information included in the middle boundary. However, the spatial variability in city (e) is not exhausted by this middle boundary; there are a total of six other frontiers of racial difference that should be considered in any spatial segregation measure. We therefore content that additive decomposability is a desideratum of nonspatial measures (it is satisfied by $I(X,Y)$), but not of explicitly spatial ones. 

@@

	\begin{figure}
		\centering
		\includegraphics[width=.5\textwidth]{figs/decomposability.png}
		\caption{A problem with additive spatial decomposability: aggregating the inner two regions transforms (e) into (f), smoothing out all spatial variability. A ``between groups'' term must therefore vanish, while a ``within-groups'' term considers only the inner boundary; not the outer ones.}
		\label{fig:decomposability}
	\end{figure}

	In summary, $J(X,Y)$ possesses a variety of useful, intuitive properties. The ones it fails to possess reflect the fact that these desiderata are typically formulated to apply to a single measure. While we have not exhausted the full complement of desiderata for segregation measures. However, many of the remaining ones -- such as composition invariance -- are either difficult to define mathematically or controversial within the sociological community, and we will not them consider further.

	We close with a discussion of some of the theoretical and operational virtues of $I(X,Y)$ and $J(X,Y)$ not considered above. 
	\begin{description}
		\item[Interpretability:] In contrast to many previous works, we make no attempt to encapsulate every aspect of diversity in a single measure. The result of maintaining distinct measures is that each of our two separate measures are highly interpretable. The mutual information $I(X,Y)$has a standard interpretation in information theory: it is the epistemic value of knowing where an individual lives for the purposes of predicting their race, as measured using the logarithmic loss function; see \cite{Cover1991} for a detailed discussion of information measures and prediction. The local information $J(X,Y)$ is then interpretable as the epistemic value of knowing where an individual lives when restricting one's attention to a small area. The direct operational meaning of these metrics is the source of their desirable mathematical properties, and reflects the virtue of starting from a unified mathematical framework.  
		\item[Kernel Independence:] Many existing spatial information measures depend on a choice of spatial kernel function which serves as a functional definition of proximity. While providing flexibility, dependence on a spatial kernel function increases researcher degrees of freedom and makes it more difficult to compare studies. There is no theoretical consensus on the appropriate kernel. The mutual information $I(X,Y)$ is aspatial and requires no kernel. In practice, computing $J(X,Y)$ does require a choice of a grid radius (see Appendix for details), but does not require the specification of a functional kernel. 
		\item[Isotropy:] As a result of its kernel independence, $J(X,Y)$ is \emph{isotropic}, taking no consideration of natural boundaries or infrastructure. This property has both virtues and drawbacks. On the one hand, an isotropic measure is simple to compute, simple to interpret, and easily generalized. On the other, boundaries and infrastructure unquestionable shape interactions between people, and are clearly relevant in segregation studies. We note, however, that most urban anisotropies exist on the scale of a few hundred meters at most. Therefore, we expect that anisotropic measures will only be relevant at scales of analysis substantially smaller than the 500m used in this study. A detailed numerical comparison of isotropic and anisotropic measures would be most welcome. 
		\item[Connection to Data Analysis:] As we will see below, the mutual information $I(X,Y)$ can be used to define a practical algorithm for the identification of natural, demographically coherent neighborhoods, and the local information $J(X,Y)$ predicts its performance. These two measures are thus not only of theoretical interest, but also provide tools for the practical analysis of urban diversity. We are not aware of any other measures that share this helpful property. 
	\end{description}

@@

	Equation \eqref{eq:information_decomp} motivates a simple scheme for identifying natural neighborhoods based on maximizing the between-group term $I(C,Y)$. If all I had access to were the cluster labels $C$ and not the locations $X$, then the information I had about $Y$would be $I(C,Y)$. A ``good'' clustering therefore maximizes the between-cluster information $I(Y,C)$, which entails minimizing the within-cluster information $I(X,Y|C)$. Solving this problem exactly is a challenging discrete optimization problem, and may not be computationally tractable. We can, however, construct a greedy algorithm which leads to satisfactory results. At each stage of the algorithm, we consider the problem of choosing a pair of tracts $\{i*,j*\}$ to cluster together. The reduction in information associated with aggregating the locations $I$ into a single cluster is 
	\begin{align}
		d(i,j) &\triangleq  \sum_{k \in \{i,j\}} p(X = k)D[p(Y|X = k)\| p(Y|X \in \{i,j\})] \\
		&- p(X\in \{i,j\})D[p(Y|X\in \{i,j\}) \| p(Y)] \label{eq:info_dist}
	\end{align}
	where $p(Y) = \sum_{x} p(x,Y)$ is the global marginal distribution. the first term is the information associated with the two separate tracts, while the second is the information associated with a merged tract. Equation \eqref{eq:info_dist} defines a natural information distance between locations $i$ and $j$. Like the KL divergence, this distance is strictly nonnegative; unlike the KL divergence, it is symmetric, and defines an axiomatic metric on the space of tracts. Importantly, we can therefore use the distance $d(i,j)$ as a dissimilarity measure for the purposes of clustering. Our greedy procedure is simple: at each iteration, determine
	\begin{equation}
		(i^*,j^*) = \argmin_{i \text{ neighbors } j} d(i,j),
	\end{equation}
	and then combine $i^*$ amd $j^*$ into a single tract, repeating until only one cluster remains. This procedure defines a form of agglomerative hierarchical clustering distinguished by two characteristics: its spatial constraints and its pursuit of minimal information loss at each step. As a greedy algorith, it possesses no guarantees for optimal solutions, but in practice its performance leads to intuitive, racially-coherent regions. It thereby enables a study of spatial difference using non-arbitrarily-defined regions.

### 2016-08-31 tex/paper/results.tex
minor updates to slides and paper

	Figure \ref{fig:AUC} also shows the bounding rectangle, whose top right corner is defined by the number N of tracts in the census data set and the full mutual information $I(X,Y)$. The AUC is defined as the ratio of the shaded area in Figure \ref{fig:AUC} to the bounding rectangle. An AUC of 1 indicates that just one region carries full information; this can only occur in a city with no spatial variation, such as city (b) in \ref{fig:toy}. A larger AUC indicates that sample models with few regions capture more information about spatial variation in the city. In Detroit, there exists a clear dividing line between a predominantly white region on the west and a predominantly black region in the east. A two-cluster model therefore captures much of the information in the city, giving a high AUC of 0.74. In Boston, on the other hand, no such clear divide exists, and more complex models are necessary to capture similar amounts of information. Thus, Boston has a lower AUC of 0.67.

@@

	Based on our discussion so far, we would expect that the AUC is related to the information measure $J(X,Y)$. Figure \ref{fig:info_and_clusters} confirms this expectation. Overall, two factors explain much about the ``clusterability'' of a city as measured by the AUC. The most important factor is the overall mutual information $I(X,Y)$. This reflects a simple intuition: when there is little spatial variation, there is not much to cluster. The second factor is the mean local information $J(X,Y)$, reflecting that spatial complexity makes clustering harder. A simple linear regression makes these insights precise. As predictors of the AUC, both $I(X,Y)$ and $J(X,Y)$ are highly significant, and the coefficient of $J(X,Y)$ is negative. The two measures jointly account for an adjusted 70\% of the variation in AUC.

>>>>

	Figure \ref{fig:AUC} also shows the bounding rectangle, whose top right corner is defined by the number N of tracts in the census data set and the full mutual information $I(X,Y)$. The AUC is defined as the ratio of the shaded area in Figure \ref{fig:AUC} to the bounding rectangle, and is computed via the formula

	\begin{equation}
		A = \frac{1}{I(X,Y) \log N }\sum_{n = 1}^N I(C_n, Y) \log \frac{n + 1}{n}  ,
	\end{equation}

	where $N$ is the total number of Census blockgroups and $C_n$ is the cluster label at the level in which $n$ total clusters are defined.

	When $A = 1$, it must hold that $I(C_1,Y) = I(X,Y)$. This implies that just one region carries full information; this can only occur in a city with no spatial unevenness, such as city (b) in \ref{fig:toy}. More generally, a large AUC indicates that simple models with few distinct regions capture much of the information about spatial variation in the city. In Atlanta, there exists a clear dividing line between a predominantly white region in the north and a predominantly black region in the south. A two-cluster model therefore captures much of the information in the city, giving a high AUC of 0.71. In Boston, on the other hand, no such clear divide exists, and more complex models are necessary to capture similar amounts of information. Thus, Boston has a lower AUC of 0.65.

@@

	We argued previously that $H(Y)$, $I(X,Y)$, and $J(X,Y)$ provide a multi-dimensional characterization of spatial diversity in a city. Based on this discussion, we would expect that the performance of our clustering algorithm, as measured by the AUC, is closely related to these measures. Figure \ref{fig:info_and_clusters} confirms this expectation using simple linear regression of the AUC on the three measures. Overall, these three information measures explain approximately an adjusted 75\% of the ``clusterability'' of cities. The signs associated with $I(X,Y)$ and $J(X,Y)$are suggestive: it is easiest to identify natural neighborhoods when there is substantial spatial variation (high $I(X,Y)$) but relatively simple local structure (low $J(X,Y)$).
